Special Constructions in Categories

### 2017-08-27 papers/talk-6/paper-6.tex
writeup6: example pushouts in sets (and add missing \labels)

\begin{example}[Solution sets]
  Let $X, Y$ be sets and $s, t \from X \to Y$ functions.
  We define $E := \setMap{x \in X}{s(x)=t(x)}$.
  Because $E \subseteq X$ we can define the \emph{inclusion function} $i \from E \to X$ with $i(e) := e$.
  Now we show that the fork
  $ \xymatrix{
    E \ar[r]^i & X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y
  } $
  is the equalizer of $s$ and $t$. This diagram commutes by construction.
  Let $A$ be another set and
  $ \xymatrix{
    A \ar[r]^f & X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y
  } $
  be another fork.
  We then define $\fbar \from A \to E$ with $\fbar(a) = f(a)$.
  Then $i \of \fbar = f$ holds by definition.
  Let $\ftilde \from A \to E$ be another function such that $i \of \ftilde = f$.
  Then $\ftilde(a) = i (\ftilde(a)) = f(a) = i (\fbar(a)) = \fbar(a)$.
\end{example}

@@

\begin{definition}[Symmetric reflexive transitive closure]
  Let $R$ be a binary relation over a set $X$.
  $\hat{R} := R \cup R^{-1}$ is the symmetric closure of $R$ where $R^{-1}$ is defined by
  $R^{-1} := \setMap{\pair{b}{a}}{\pair{a}{b} \in R}$.
  The \demph{symmetric transitive closure} of $R$ is $\sim := \bigcup_{n\in\nat} {\hat R}^n$
  where ${\hat R}^n$ is recursively defined with
  ${\hat R}^0 := \setMap{\pair{a}{a}}{a \in X}$ and
  ${\hat R}^{n+1} := {\hat R}^n \of \hat R = \setMap{\pair{a}{c}}{\pair{a}{b} \in {\hat R}^n \land \pair{b}{c} \in \hat R}$.
  It can be shown that $\sim$ is the \emph{least equivalence relation} containing $R$.
\end{definition}

\begin{example}[Quotient sets]
  Let $X, Y$ be sets and $s, t \from X \to Y$ be functions.
  We have to give a set $C$ and a function $i$ such that the diagram
  $ \xymatrix{
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[r]^i & C
  } $
  commutes.
  Therefore we define the relation
  $R := \setMap{\pair{s(a)}{t(a)}}{a \in Y} \subseteq Y^2$.
  Let $\sim \; \subseteq Y^2$ be the symmetric reflexive transitive closure of $R$.
  Now we define $C := Y / \sim = \setMap{\class y}{y \in Y}$ where $\class y := \setMap{y'}{y \sim y'}$ are the equivalence classes.
  For the function $i \from Y \to E$ we choose $i(y) := \class y$.
  Then the diagram
  $ \xymatrix{
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[r]^i & C
  } $
  commutes because for $x \in X$, $s(x)$ and $t(x)$ are in $R$ and therefore also in $\sim$; because $s(x) \sim t(x)$ we have $i(s(x)) = \class{s(x)} = \class{t(x)} = i(t(x))$.
  Now let
  $ \xymatrix{
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[r]^f & A
  } $
  be a cofork.
  By induction we can show that $f$ ``does not care'' about equal values, i.~e. $f(y_1) = f(y_2)$ whenever $y_1 \sim y_2$.

@@

  Now we can define $\fbar \from C \to A$ by $\fbar(e) := f(\choice e)$.
  We now have  $\fbar(i(y)) = f(\choice{\class y}) = f(y)$ because $y \sim \choice{\class y}$ (by the choice axiom and reflexivity of $\sim$).
  Now we only have to show that this $\fbar$ is unique in respect to that property.
  So we assume another function $\ftilde \from C \to A$ with $\ftilde \of i = f$. But then we have
  $$ \fbar(c) \explainRel{Def}= f(\choice c) \explainRel{Ass.}= \ftilde(i(\choice c)) \explainRel{Def}= \ftilde(\class{\choice c}) \explainRel{Choice}= \ftilde(c).$$
\end{example}

@@

\begin{example}[Pullbacks and products]
  When the pullback square
  \[ \xymatrix{
    P \ar[r]_{p_2} \ar[d]^{p_1} & Y \ar[d]_{t} \\
    X \ar[r]^s & Z
  } \]
  is the pullback of
  $ \xymatrix{
    X \ar[r]^s & Z & Y\ar[l]_t \\
  } $
  and $Z$ is a terminal object then $P$ is the product of $X$ and $Y$.
\end{example}

\begin{example}[Pullbacks in sets]
  Let $X, Y, Z$ be sets and $s \from X \to Z, t \from Y \to Z$ be functions.
  Then $P := \setMap{\pair{x}{y}}{s(x) = t(y)}$ with the projection functions
  $p_1(x,y) := x$ and $p_2(x, y) := y$ is the pullback of
  $ \xymatrix{
    X \ar[r]^s & Z & Y\ar[l]_t \\
  } $.
  The pullback square commutes by construction.
  Now let $A$ be a set and $f_1 \from A \to X, f_2 \from A \to Y$ be functions with $s \of f_1 = t \of f_2$.
  Then we define $\fbar \from A \to P$ by
  $\fbar(a) := \pair{f_1(a)}{f_2(a)}$.
  This is well defined because for $a \in A$ we have $\fbar(a) = \pair{f_1(a)}{f_2(a)} \in P \Leftrightarrow s(f_1(a)) = t(f_2(a))$ which is true by assumption.
  We also have $p_1(\fbar(a)) = f_1(a)$ and $p_2(\fbar(a)) = f_2(a)$ by constructions.
  We also see that $\fbar$ is unique in respect to that property.
\end{example}

@@

\todo{Beispiele}

\section*{Conclusion}
We have seen much analogous definitions.
Our definitions follow a simple ``pattern'':
\todo{Pattern könnte besser erklärt werden aber nicht zu formal!}
\begin{itemize}
  \item First we have some objects or a commuting diagram ($D_0$).
  \item Then we ``insert'' an object $P$ and get a new commuting diagram ($D_1$) where we ``connect'' $P$ with the objects of $D_0$ with ``projection functions''.
  \item Then we ``replace'' this object with a new object, called $A$, and replace the ``projection functions'' with new functions. We get a new diagram ($D_2$).
  \item For each $A$ there exists a unique map $\fbar$ such if we ``merge'' $D_1$ and $D_2$ we get a new commuting diagram $D_3$, where $A$ and $P$ are connected with $\fbar$.
\end{itemize}

>>>>

\begin{example}[Solution sets]
  \label{ex:equalizer:solset}
  Let $X, Y$ be sets and $s, t \from X \to Y$ functions.
  We define $E := \setMap{x \in X}{s(x)=t(x)}$.
  Because $E \subseteq X$ we can define the \emph{inclusion function} $i \from E \to X$ with $i(e) := e$.
  Now we show that the fork
  $ \xymatrix{
    E \ar[r]^i & X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y
  } $
  is the equalizer of $s$ and $t$. This diagram commutes by construction.
  Let $A$ be another set and
  $ \xymatrix{
    A \ar[r]^f & X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y
  } $
  be another fork.
  We then define $\fbar \from A \to E$ with $\fbar(a) = f(a)$.
  Then $i \of \fbar = f$ holds by definition.
  Let $\ftilde \from A \to E$ be another function such that $i \of \ftilde = f$.
  Then $\ftilde(a) = i (\ftilde(a)) = f(a) = i (\fbar(a)) = \fbar(a)$.
\end{example}

@@

\begin{definition}[Symmetric reflexive transitive closure]
  \label{def:srtc}
  Let $R$ be a binary relation over a set $X$.
  $\hat{R} := R \cup R^{-1}$ is the symmetric closure of $R$ where $R^{-1}$ is defined by
  $R^{-1} := \setMap{\pair{b}{a}}{\pair{a}{b} \in R}$.
  The \demph{symmetric transitive closure} of $R$ is $\sim := \bigcup_{n\in\nat} {\hat R}^n$
  where ${\hat R}^n$ is recursively defined with
  ${\hat R}^0 := \setMap{\pair{a}{a}}{a \in X}$ and
  ${\hat R}^{n+1} := {\hat R}^n \of \hat R = \setMap{\pair{a}{c}}{\pair{a}{b} \in {\hat R}^n \land \pair{b}{c} \in \hat R}$.
  It can be shown that $\sim$ is the \emph{least equivalence relation} containing $R$.
\end{definition}

\begin{example}[Quotient sets]
  \label{ex:coequa:quotient}
  Let $X, Y$ be sets and $s, t \from X \to Y$ be functions.
  We have to give a set $C$ and a function $i$ such that the diagram
  $ \xymatrix{
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[r]^i & C
  } $
  commutes.
  Therefore we define the relation
  $R := \setMap{\pair{s(a)}{t(a)}}{a \in Y} \subseteq Y^2$.
  Let $\sim \; \subseteq Y^2$ be the symmetric reflexive transitive closure of $R$.
  Now we define $C := Y / \sim = \setMap{\class y}{y \in Y}$ where $\class y := \setMap{y'}{y \sim y'}$ are the equivalence classes.
  For the function $i \from Y \to E$ we choose $i(y) := \class y$.
  Then the diagram
  $ \xymatrix{
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[r]^i & C
  } $
  commutes because for $x \in X$, $s(x)$ and $t(x)$ are in $R$ and therefore also in $\sim$; because $s(x) \sim t(x)$ we have $i(s(x)) = \class{s(x)} = \class{t(x)} = i(t(x))$.
  Now let
  $ \xymatrix{
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[r]^f & A
  } $
  be a cofork.
  By induction we can show that $f$ ``does not care'' about equal values, i.~e. $f(y_1) = f(y_2)$ whenever $y_1 \sim y_2$.

@@

  Now we can define $\fbar \from C \to A$ by $\fbar(e) := f(\choice e)$.
  We now have  $\fbar(i(y)) \explainRel{def}= f(\choice{\class y}) = f(y)$ because $y \sim \choice{\class y}$ (by the choice axiom and reflexivity of $\sim$).
  Now we only have to show that this $\fbar$ is unique in respect to that property.
  So we assume another function $\ftilde \from C \to A$ with $\ftilde \of i = f$. But then we have
  $$ \fbar(c) \explainRel{def}= f(\choice c) \explainRel{ass}= \ftilde(i(\choice c)) \explainRel{def}= \ftilde(\class{\choice c}) \explainRel{choice}= \ftilde(c).$$
\end{example}

@@

\begin{example}[Pullbacks and products]
  \label{ex:pullback:product}
  When the pullback square
  \[ \xymatrix{
    P \ar[r]_{p_2} \ar[d]^{p_1} & Y \ar[d]_{t} \\
    X \ar[r]^s & Z
  } \]
  is the pullback of
  $ \xymatrix{
    X \ar[r]^s & Z & Y\ar[l]_t \\
  } $
  and $Z$ is a terminal object then $P$ is the product of $X$ and $Y$.
\end{example}

\begin{example}[Pullbacks in sets]
  \label{ex:pullback:set}
  Let $X, Y, Z$ be sets and $s \from X \to Z, t \from Y \to Z$ be functions.
  Then $P := \setMap{\pair{x}{y}}{s(x) = t(y)}$ with the projection functions
  $p_1(x,y) := x$ and $p_2(x, y) := y$ is the pullback of
  $ \xymatrix{
    X \ar[r]^s & Z & Y\ar[l]_t \\
  } $.
  The pullback square commutes by construction.
  Now let $A$ be a set and $f_1 \from A \to X, f_2 \from A \to Y$ be functions with $s \of f_1 = t \of f_2$.
  Then we define $\fbar \from A \to P$ by
  $\fbar(a) := \pair{f_1(a)}{f_2(a)}$.
  This is well defined because for $a \in A$ we have $\fbar(a) = \pair{f_1(a)}{f_2(a)} \in P \Leftrightarrow s(f_1(a)) = t(f_2(a))$ which is true by assumption.
  We also have $p_1(\fbar(a)) = f_1(a)$ and $p_2(\fbar(a)) = f_2(a)$ by constructions.
  We also see that $\fbar$ is unique in respect to that property.
\end{example}

@@

\begin{example}[Pushouts in sets]
  \label{ex:pushout:set}
  To give an example for pushouts in sets we somehow have to combine the coproduct (disjoint union; see example \ref{ex:coprod:dunion}) and the coequalizer (quotient set; see example \ref{ex:coequa:quotient}).
  Let
  $ \xymatrix{
    X & Z \ar[l]_s \ar[r]^t & Y
  } $
  be a functions and sets.
  We want to define a set $P$ and two functions $f_1 \from X \to A, f_2 \from Y \to A$, such that the diagram
  \[ \xymatrix{
    Z \ar[r]_s \ar[d]^t & X \ar[d]_{p_1} \\
    Y \ar[r]^{p_2} & P
  } \]
  commutes.
  We define $P := (X \dunion Y) / \sim$ where $\sim \; \subseteq (X \dunion Y)^2$ is the symmetric reflexive transitive closure (see definition \ref{def:srtc}) over the relation
  $R := \setMap{\pair{\pair{s(z)}{0}}{\pair{t(z)}{1}}}{z \in Z} \subseteq (X \dunion Y)^2$.
  When we choose the projection functions
  \begin{alignat*}{2}
    p_1(x) &:= \class{\pair{x}{0}} \\
    p_2(y) &:= \class{\pair{y}{1}}
  \end{alignat*}
  the diagram commutes by construction. Now let
  \[ \xymatrix{
    Z \ar[r]_s \ar[d]^t & X \ar[d]_{f_1} \\
    Y \ar[r]^{f_2} & A
  } \]
  be another pushout square. To define the function $\fbar \from P \to A$ we first define a helper function $\fbar' \from X \dunion Y \to A$.
  Therefore we use the choice function
  $\choice \cdot \from P \to X \dunion Y$:
  \begin{alignat*}{2}
    \fbar'(x, 0) &:= f_1(x) \\
    \fbar'(y, 1) &:= f_2(y) \\
    \fbar(p)     &:= \fbar'(\choice p)
  \end{alignat*}
  By induction we can show that $\fbar'$ ``does not care'' about equal values, i.~e. $\forall a,b \in X \dunion Y, a \sim b \Rightarrow \fbar'(a) = \fbar'(b)$.

  We show $f_1 = \fbar \of p_1$; $f_2 = \fbar \of p_2$ follows analogously.
  $$\fbar(p_1(x)) \explainRel{def}= \fbar'(\choice {\class {\pair x 0}}) = \fbar'(x, 0) \explainRel{def}= f_1(x).$$
  The second step follows by $\pair x 0 \sim \choice{\class {\pair x 0}}$ which holds by the choice axiom and reflexivity of $\sim$.

  The last step is again to show that $\fbar$ is unique in respect to that property. So we assume another function $\ftilde \from X \dunion Y \to A$ with $f_1 = \ftilde \of p_1$ and $f_2 = \ftilde \of p_2$.
  Let $p \in P$. We assume without loss of generality $\choice p = \pair x 0$ for some $x \in X$. Then
  $$\fbar(p)
    \explainRel{def}    = \fbar'(\choice p)
    \explainRel{ass}    = \fbar'(x, 0)
    \explainRel{def}    = f_1(x)
    \explainRel{ass}    = \ftilde(p_1(x)) \explainRel{def}= \ftilde(\class{\pair x 0})
    \explainRel{ass}    = \ftilde(\class{\choice p})
    \explainRel{choice} = \ftilde(p).$$
\end{example}

\section*{Conclusion}
We have seen much analogous definitions and proofs.
Our definitions follow a simple ``pattern'':
\todo{Pattern könnte besser erklärt werden aber nicht zu formal!}
\begin{itemize}
  \item First we have some objects or a commuting diagram ($D_0$).
  \item Then we ``insert'' an object $P$ and get a new commuting diagram ($D_1$) where we ``connect'' $P$ with the objects of $D_0$ with ``projection functions''.
  \item Then we ``replace'' this object with a new object, called $A$, and replace the ``projection functions'' with new functions. We get a new diagram ($D_2$).
  \item For each $A$ there exists a unique map $\fbar$ such if we ``merge'' $D_1$ and $D_2$ we get a new commuting diagram $D_3$, where $A$ and $P$ are connected with $\fbar$.
\end{itemize}
